---
title: "check_de_sample_outliers"
author: "Jared Slosberg"
date: "3/27/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(monocle3)
library(tidyverse)
library(here)
library(DESeq2)
```

```{r}

#load deseq object
dds <- readRDS(here("data/bulkrna_dds.rds"))
metadata <- read.csv(here("metadata_merged_filtered.csv"))

#load de gene modules
de_module_df <- read.csv(here("results/dynamic_gene_modules.csv"))
de_genes <- de_module_df %>% pull(ensemblgene_id)

```

```{r, eval = F}
#sample specific outliers make up most of clusters 5 & 7. These are the easiest to see, but "rest_genes" also includes some of these
suspect_genes <- de_module_df %>% filter(cluster.k.7 %in% c(5,7)) %>% pull(ensemblgene_id)
suspect_counts <- assay(dds[suspect_genes,])
head(suspect_counts)
quantile(suspect_counts, seq(from = 0, to = 1, by = .1))

rest_genes <- de_module_df %>% filter(!(cluster.k.7 %in% c(5,7))) %>% pull(ensemblgene_id)
rest_counts <- assay(dds[rest_genes,])
head(rest_counts)
quantile(rest_counts, seq(from = 0, to = 1, by = .1))

#count matrix for suspect genes is 90% zeros compared to <50% for the rest of de genes
```

How many samples are de genes detected in (counts > 0)
```{r}
n_samples <- ncol(dds)
n_vec <- 1:n_samples

pctNonzeroRows <- function(mat){
  pct_nonzero <- apply(mat, MARGIN = 1, FUN = function(x){
    sum(x > 0)/length(x)
  })
  
  return(pct_nonzero)
}

#set up gene list to map over
gene_list <- list("all" = de_genes)

pct_pass_df <- map_df(names(gene_list), function(gene_class){

  genes <- gene_list[[gene_class]]
  pct_df <- map_df(n_vec, function(i){
    thresh <- i/n_samples
    
    #for each gene, how many samples is it detected in
    gene_nonzero_pct <- assay(dds[genes,]) %>%
      pctNonzeroRows()
    
    #for all genes, at this threshold, how many genes meet or pass it
    pct_genes_pass <- sum(gene_nonzero_pct >= thresh)/length(gene_nonzero_pct)
    
    df <- data.frame(samp_threshold = as.factor(i), pct_genes_pass = pct_genes_pass, class = gene_class)
    
  })
})

ggplot(pct_pass_df) + geom_bar(aes(x = samp_threshold, y= pct_genes_pass, fill = class), stat = "identity", position = "dodge") +
  xlab("Number of samples detected in") + 
  ylab("Proportion of genes")


# %>%
#   full_join(de_module_df) %>%
#   mutate(class = case_when(
#     cluster.k.7 %in% c(5,7) ~ "suspect",
#     !(cluster.k.7 %in% c(5,7)) ~ "not_suspect"
#     
#   ))
```


```{r}
test_names <- paste0("age_",c("6mo","17mo"),"_vs_P30")

pcut <- 0.05
clip_nlog10p <- 10
sigthresh_nlog10p <- -log10(.05)

res_df <- map_df(test_names, function(name){
  tbl <- read.csv(paste0(here("results/diff_exp/"), name,".csv"), row.names = 1)
  tbl <- tbl %>% mutate(condition = name,
                        nlog10p = ifelse(-log10(padj) > clip_nlog10p, clip_nlog10p, -log10(padj)),
                        nlog10p_clipped = ifelse(-log10(padj) > clip_nlog10p, T, F ))
  return(tbl)
})

res_df_filtered <- res_df %>% filter(ensemblgene_id %in% de_genes)
res_df_filtered_wide <- res_df_filtered  %>%
  dplyr::select(c(ensemblgene_id, padj, nlog10p, condition, log2FoldChange)) %>%
  pivot_wider(names_from = condition, values_from = c(padj, log2FoldChange, nlog10p)) %>% 
  mutate(both_sig = padj_age_6mo_vs_P30 < pcut & padj_age_17mo_vs_P30<pcut,
         doe_17mo_vs_P30 = sign(log2FoldChange_age_17mo_vs_P30),
         doe_6mo_vs_P30 = sign(log2FoldChange_age_6mo_vs_P30),
         direction_consistent = (doe_17mo_vs_P30 == doe_6mo_vs_P30),
         direction = ifelse(direction_consistent, sign(log2FoldChange_age_6mo_vs_P30), paste(doe_6mo_vs_P30,doe_17mo_vs_P30, sep = ","))) 

log2fc_long <- res_df_filtered %>%
  full_join(de_module_df) %>%
  mutate(condition = 
           case_when(condition == "age_6mo_vs_P30" ~ "6mo", condition == "age_17mo_vs_P30" ~ "17mo")) %>%
  dplyr::select(-c(baseMean, lfcSE, stat, pvalue, padj, nlog10p, nlog10p_clipped))

```

```{r}
clip_lfc <- 2

res_df <- res_df %>% mutate(log2FoldChange_clipped = ifelse(abs(log2FoldChange) > abs(clip_lfc), sign(log2FoldChange)*clip_lfc, log2FoldChange))
ggplot(res_df) + geom_point(aes(x = baseMean, y = log2FoldChange_clipped),color = "grey", alpha = 0.5) + scale_x_log10() + facet_wrap(~condition)

#de_genes
all_dummy <- res_df %>% mutate(gene_class = "all")
de_dummy <- res_df %>% filter(ensemblgene_id %in% de_genes) %>% mutate(gene_class = "de_genes")

dummy_df <- rbind(all_dummy, de_dummy)
ggplot(dummy_df) + ggpointdensity::geom_pointdensity(aes(x = baseMean, y = log2FoldChange_clipped), alpha = 0.5) + scale_x_log10() + facet_grid(rows = vars(gene_class), cols = vars(condition))

```

```{r}

#genes expressed at all
gene_filter <- (rowSums(counts(dds)) >= 10) %>% enframe(name = "ensemblgene_id", value = "is_expressed")

#genes expressed in 3+ samples
gene_sample_filter <- (rowSums(counts(dds) >= 10) >= 3) %>% enframe(name= "ensemblgene_id", value = "is_expressed_3_samples")

dummy_df <- dummy_df %>% left_join(gene_filter) %>% left_join(gene_sample_filter)

ggplot(dummy_df) + geom_point(aes(x = baseMean, y = log2FoldChange_clipped, color = is_expressed), alpha = 0.5) + scale_x_log10() + facet_grid(rows = vars(gene_class), cols = vars(condition))

ggplot(dummy_df) + geom_point(aes(x = baseMean, y = log2FoldChange_clipped, color = is_expressed_3_samples), alpha = 0.5) + scale_x_log10() + facet_grid(rows = vars(gene_class), cols = vars(condition))

#all
ggplot(filter(dummy_df, gene_class == "all")) + geom_point(aes(x = baseMean, y = log2FoldChange_clipped, color = is_expressed_3_samples), alpha = 0.5) + scale_x_log10() + facet_wrap(~is_expressed_3_samples, ncol = 1) + ggtitle("All genes")
ggplot(filter(dummy_df, gene_class == "de_genes")) + geom_point(aes(x = baseMean, y = log2FoldChange_clipped, color = is_expressed_3_samples), alpha = 0.5) + scale_x_log10() + facet_wrap(~is_expressed_3_samples, ncol = 1)+ ggtitle("de genes")


```
What proportion of DE genes are expressed in 3+ samples
```{r}
dummy_df %>%
  mutate(called_de = (ensemblgene_id %in% de_genes))%>%
  dplyr::select(ensemblgene_id, called_de, is_expressed_3_samples) %>%
  unique %>%
  dplyr::select(-c(ensemblgene_id)) %>% 
  table
```

Cooks distance: "The DESeq function calculates, for every gene and for every sample, a diagnostic test for outliers called Cook’s distance. Cook’s distance is a measure of how much a single sample is influencing the fitted coefficients for a gene, and a large value of Cook’s distance is intended to indicate an outlier count. "
```{r}
cutoff <- qf(0.99, 6, 10) #quantile, number of parameters estimated, number of samples - number of parameters

cooks_df <- assay(dds, "cooks")
cooks_gene_ind <- apply(as.matrix(cooks_df), MARGIN = 1, function(x){ any(x > cutoff)})
cooks_gene_ind %>% table

#how many DE genes would have been filtered out (by default remove cooks > cutoff)
assay(dds, "cooks")[de_genes,] %>% as.matrix() %>% apply(MARGIN = 1, function(x){ any(x > cutoff)}) %>% table()


par(mar=c(8,5,2,2))
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)

cooks_long <- cooks_df %>% as.data.frame() %>%
  rownames_to_column(var = "ensemblgene_id") %>%
  pivot_longer(!ensemblgene_id, names_to = "sample_name", values_to = "cooks_distance")

  # %>% mutate(outlier = ifelse(sample_name == "17mo-M-1", "17mo-M-1", NA)) #outlier

ggplot(cooks_long) + geom_density(aes(x = cooks_distance, group = sample_name)) + scale_x_log10() + geom_vline(xintercept = cutoff, color = "red")
ggplot(cooks_long) + geom_density(aes(x = cooks_distance, group = sample_name)) + geom_vline(xintercept = cutoff, color = "red")  +scale_x_log10() + coord_cartesian(xlim = c(1e-9, 1500))

```
~ 1% genes have at least one outlier sample
~ 2.5% of de genes have at least one outlier sample


By sample, what is the percentage of combos that are outliers?
- ~ 0.1% of measurements are deemed outliers
```{r}
cooks_long <- cooks_long %>% mutate(is_outlier = cooks_distance > cutoff)

outlier_df <- cooks_long %>% group_by(sample_name) %>% summarise(outlier_ratio = sum(is_outlier)/length(is_outlier))
outlier_df

```

